What is an Ancestral Recombination Graph?

### 2022-02-18 paper.tex
Add separate simplification section

\section*{Ancestry resolution and simplification}
% This is a rough first pass. Will clarify terminology later.
The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

@@

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
\caption{\label{fig-recombination-nodes}
Recombination nodes are optional in an edge-annotated ARG. (a) A simulated
ARG including all recombination nodes (red). (b) The same ARG having removed both
recombination nodes and nodes of common ancestors in which no genetic
lineages coalesce (cyan). (c) The sequence of trees corresponding to (b): if we were
to plot out the trees corresponding to (a) we would have more trees, but the topologies
and MRCA nodes at each genomic location would be identical. For clarity, nodes are
plotted by ranked time rather than true time on the y-axis.
% HYW: note that we also truncate nodes for portions in which they are unary,
% so that nodes which are coalescent nodes over only some of their length
% (blue-green) become coalescent nodes over their entire length (green). This
% can, however, throw away some useful information, for example node 4 has
% only a single parent edge in (a), but 2 parent edges (one of which appears
% to bypass node 11) in (b). In fact, this edge should go through node 11, but
% there may be no way to detect this from real data.
}
\end{figure}

% Not sure where this text will end up, but we want it somewhere
% towards the end.

>>>>

\section*{Ancestry resolution}
% This is a rough first pass. Will clarify terminology later.
The definition of an ARG as a data structure in terms of
individuals' monoploid genomes
and their direct patterns of inheritance allows us to describe arbitrarily
complex relatedness structures, free from the limitations imposed by
coalescent approximations.
An ARG is an intrinsically retrospective
structure, defined with respect to the ancestry of
a set of sampled genomes.
The graph structure (the nodes and their connectivity through edges)
captures a great deal of information about this ancestry,
but on its own it cannot tell us about the precise relationships between
the samples along the genome: do to this, we must construct the
local (or ``marginal'') trees.

@@

\section*{ARG simplification}

Not all nodes in an ARG are equal. Assuming that we have sequenced the
genomes of the samples, there are intrinsic limits on what can
be known about more ancient nodes, determined by the graph topology
and the flow ancestral material through it.
\emph{Simplification}~\citep{kelleher2018efficient} is the process
of removing nodes and re-writing edges in a GARG to remove
various types of redundancy, while maintaining the crucial properties
that the meaningful topology in the local trees is not affected,
and the relationship between nodes in adjacent trees is maintained.

We may have nodes corresponding to genomes that are ancestral to
the sample along their entire span
(e.g.\ node $H$ in Fig.~\ref{fig-ancestry-resolution}),
genomes that carry no genetic material ancestral to the sample at all
(e.g.\ node $J$ in Fig.~\ref{fig-ancestry-resolution}),
and every possibility along the continuum between these extremes.
Clearly, we can only ever hope to infer the fraction of an ancestor's
genome that is ancestral to the sample.

\begin{figure}
\centering
\vspace{5em}
\includegraphics[width=\linewidth]{illustrations/ARG_recomb_node_deletion}
\caption{\label{fig-simplification}
ARG simplification. (a) A simulated
ARG including all recombination nodes (red). (b) The same ARG having removed both
recombination nodes and nodes of common ancestors in which no genetic
lineages coalesce (cyan). (c) The sequence of trees corresponding to (b): if we were
to plot out the trees corresponding to (a) we would have more trees, but the topologies
and MRCA nodes at each genomic location would be identical. For clarity, nodes are
plotted by ranked time rather than true time on the y-axis.
% HYW: note that we also truncate nodes for portions in which they are unary,
% so that nodes which are coalescent nodes over only some of their length
% (blue-green) become coalescent nodes over their entire length (green). This
% can, however, throw away some useful information, for example node 4 has
% only a single parent edge in (a), but 2 parent edges (one of which appears
% to bypass node 11) in (b). In fact, this edge should go through node 11, but
% there may be no way to detect this from real data.
}
\end{figure}
